---
title: "Practical 3. Binomial and Poisson examples"
author: "Conor Cuddihy"
date: "2020-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Calculating probabilities using the binomial distribution

## Galway rain example

The probaility that it rains in Galway on a given day is 0.8. You are interested in $X$ the number of rainy days in the next week. 

What is the probability that it rains every day?
 
This can be answered by assuming that the random variable of interest (i.e. the number of rainy days) follows a binomial distribution with probability 0.8 and number of trials equal to 3, i.e. $X\sim Binomial(n=7,p=0.8)$. The command needed is `dbinom(x, size, prob)`.
 

```{r}
dbinom(x=7, size=7, prob=0.8)
```

1. Find the probability that it rains twice in the next week


```{r}

```

2. Find the probability that it rains less than four times in the next week

```{r}

```


3. Find the probability that it rains between 2 and 4 times in the next week, i.e. $2 \leq X \leq 4$

```{r}

```






## Pollutant example

You are working for the EPA and are interested in the levels of water pollution in Galway, where previous estimates suggest 10% of samples are polluted. Let $X$ denote the number of samples that contain pollutant in the next 18 samples analyzed. Then $X$ is a binomial random variable with $p = 0.1$ and $n = 18$. 

4. Find the probability that, in the next 18 samples exactly 2 contain the pollutant.


```{r}

```


5. Determine the probability that there are between 3 and 6 samples contain pollutants i.e. $3 \leq X \leq 6$. 

```{r}

```


# Calculating probabilities using the Poisson distribution

## Burglaries in Galway

There was a [30% increase in monthly burglaries in county Galway in 2017](https://www.galwaydaily.com/news/huge-increase-in-burglary-rate-across-galway/), with a total of 672 (1.84 per day) reported to the Gardai. Suppose the number of burglaries in Galway follows a Poisson distribution with mean of 1.84 per day.

6. Find the probability that there will be exactly 3 burglaries in Galway in one day
 
```{r}

```

7. Find the probability that there will be less than 3 burglaries in Galway in one day
 
```{r}

```

8. Find the probability that there will be more than 20 burglaries in a week
 
```{r}

```


## Flaws on Copper Wire example.

Suppose that the number of flaws on a thin copper wire follows a Poisson distribution with a mean of 2.3 flaws per mm. 

9. Find the probability of exactly 2 flaws in 1 mm of wire
 
 
```{r}

```

10. Determine the probability of 10 flaws in 5 mm of wire.

```{r}

```
 
11. Find the probability that there will be more than 10 flaws in 5mm of wire.

```{r}

```


# Calculating probabilities using the Normal distribution

Suppose the commute to work in NUI, Galway follows a Normal distribution with mean 20 and standard deviation 5. 

12. Find the probability of making it to work in 10 minutes or less

```{r}

```

13. Find the probabaility that it takess more than 25 minutes to get to work

```{r}

```


14. Find the probability of making it to work in between 10 and 15 minutes

```{r}

```

15. Determine the time below which 90% of commutes are lower than

```{r}

```