---
title: "Practical 4. Training Intervention Analysis"
author: "Conor Cuddihy"
date: "2020-10-04"
output: word_document
---

## Context: Celtic Study introduced in class.

Scaffold for the analysis when the primary response variable is VO2 max.  You need to rerun the analysis using the Squat variables (i.e. Squat_Pre, Squat_Post) to see if there has been any improvement on average and provide a tolerance interval for the likely improvement for 95% of players in the population of interest (with 95% confidence).


```{r}
library(infer)
library(tidyverse)
library(tolerance)
```

## Read in the training intervention data

Read in the data and have a look at the variable names and structure of the data.

```{r}

train.df <- read.csv("training_intervention.csv")
glimpse(train.df)
```

## Focus on the VO2 max response variables.


## Summary Statistics of Pre and Post

1. Summarise the pre- and post-intervention VO2 max 


```{r}
train.df %>% select(VO2.max_Pre,VO2.max_Post) %>% summary()

```

## Mean and Standard Deviation of Pre and Post

2. Get the mean and standard deviation of pre- and post-intervention 

```{r}

train.df %>% select(VO2.max_Pre,VO2.max_Post) %>%
            summarize(Pre_Mean=mean(VO2.max_Pre), Pre_SD= sd(VO2.max_Pre),
                      Post_Mean=mean(VO2.max_Post), Post_SD= sd(VO2.max_Post))

```

## Scatterplot of Pre and Post with line of equality

3. Produce a scatterplot of pre- and post-intervention VO2 max along with the line of equality 


```{r}
train.df %>% ggplot(aes(x = VO2.max_Pre, y = VO2.max_Post)) +
        geom_point() + 
  ggtitle("Scatterplot of Pre and Post VO2 Max") +
  ylab("Post VO2 Max (units ?)") +
  xlab("Post VO2 Max (units ?)") +
  geom_abline(slope=1, intercept=0)
  
```

4. Interpret the plot in your own words



## Calculate the Improvement

5. Calculate a new variable and have a look at the data frame to see that it has been created.  High values of VO2 max are good to Post-Pre is a better measure than Pre-Post to capture this.

```{r}

train.df <- train.df %>% mutate(Improvement = VO2.max_Post-VO2.max_Pre) %>%
              glimpse()
  


```

## Mean and Standard Deviation of Improvement

6. Calculate the mean and SD of Improvement

```{r}

train.df %>% select(Improvement) %>%
            summarize(Imp_Mean=mean(Improvement), Imp_SD= sd(Improvement))

```


## Boxplot of Improvement

7. Produce a boxplot of Improvement

```{r}


train.df %>% ggplot(aes(x = "", y = Improvement)) +
        geom_boxplot() + 
  ggtitle("Boxplot of Improvement in VO2 Max") +
  ylab("Improvement in VO2 Max (units ?)") +
  xlab("") +
  coord_flip()

```

## 95% Confidence Interval for the mean

8. Calculate and interpret a 95% confidence interval for the population mean improvement in VO2 max (hint: use the t.test function)

```{r}

train.df %>% select(Improvement) %>% t.test()

```


## 95% Bootstrap CI for the mean

9. Calculate a 95% bootstrap interval for th mean improvement in VO2 max

```{r}

boot <- train.df %>%
  specify(response = Improvement) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "mean")

percentile_ci <- get_ci(boot)
round(percentile_ci,2)

```


10. Visualise the bootstrap interval and comment on any differences with the 95% confidence interval above

```{r}
boot %>% visualize(endpoints = percentile_ci, direction = "between") +
                   xlab("Bootstrap Mean") + ylab("Frequency")

```


## 95% Tolerance Interval

11. Calculate a 95% tolerance interval covering 95% of improvement values 

```{r}

normtol.int(train.df$Improvement, alpha = 0.05, P = 0.95)

```


## Conclusion

12. Write up a conclusion using the plots and intervals you have obtained above
