---
title: "Practical 6. Compressive strength of concrete"
author: "Conor Cuddihy"
date: "2020-10-17"
output: html_document
---

An engineer wishes to compare the mean compressive strength of concrete from two plants. A random sample of 28 batches of concrete from plant A and a random sample of 72 batches from plant B were tested and the compressive strength (mega pascals, MPa) of each recorded.  

Create relevant plots and analyse the data appropriately to decide if there evidence of a significant difference in the mean compressive strength of concrete between the two plants.



```{r}
library(infer)
library(tidyverse)
library(tolerance)
```



```{r}
strength_df <- read.csv("strength.csv")
glimpse(strength_df)
```

## Summary Statistics


1. Calculate the mean and standard deviation of strength for each plant separately

```{r}
strength_df %>% group_by(plant) %>% summarise(n=n(),
                                              mean=mean(Compressive.Strength),
                                              sd=sd(Compressive.Strength))
```

## Boxplot

2. Create side-by-side boxplots of strength by plant and **interpret these plots**

```{r}
ggplot(strength_df, aes(x = plant, y = Compressive.Strength, fill = plant)) + 
  geom_boxplot() + labs(y = "Compressive Strength (MPa)")

```


## t-test and 95% Confidence Interval for difference in the population mean

3. Carry out a two sample t-test to compare strength of concrete from plants A and B

```{r}
# t test for comparing type A and B
strength_df %>%
t_test(Compressive.Strength ~ plant, order = c("B", "A"), 
       alternative = "two_sided",
       mu=0,
       conf_level = 0.95,
       paired=FALSE,
       var.equal=FALSE)

```

4. What are the null and alternative hypotheses for this test? 

5. Use the p-value to decide whether to reject the null hypothesis or not

6. Interpret the 95% confidence interval for the mean difference

## 95% Bootstrap CI for difference in means

7. Use the bootstrap to obtain a 95% confidence interval for the population mean difference

8. Why is this interval not exactly the same as the confidence interval?

```{r}

strength_boot <- strength_df %>%
  specify(response =Compressive.Strength, explanatory = plant) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c("B", "A"))

get_ci(strength_boot)

strength_boot %>% visualize(endpoints = get_ci(strength_boot), direction = "between")

```

## Tolerance Intervals

9. Obtain 95%/95% tolerance intervals for the compressive strength of concrete from plants A and B separately and **interpret** these intervals carefully.

### Type A

```{r}

normtol.int(strength_df$Compressive.Strength[strength_df$plant=="A"], alpha = 0.05, P = 0.95, side = 2)

```

### Type B

```{r}

normtol.int(strength_df$Compressive.Strength[strength_df$plant=="B"], alpha = 0.05, P = 0.95, side = 2)

```

## Conclusion

10. Write an overall conclusion and decision as to which plant to purchase from

