---
title: "Practical 7. Correlation and Regression"
author: "Conor Cuddihy"
date: "2020-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Modelling the effect of age on Netflix browsing

###Study Description: 
A psychologist is interested in how long people spend on Netflix per day and has collected data from $n=93$ accounts belonging to NUI Galway staff and students. You've been asked to lend statistical support to this project and want to implement a regression model to find out whether age is a useful predictor of 'Netflix time'. Let $Y$ be the time in minutes that is spent on Netflix and assume that this is Normally distributed. Let $X$ be the age of students with data available.

### Aims: 
To investigate whether there is sufficient evidence of a dependency of Netflix time on age in the population of NUI Galway staff and students.



```{r}
# Load the libraries needed. You may need to install moderndive 
library(tidyverse)
library(moderndive)
```


```{r}
# read in the data
netflix_df <- read.csv("netflix_data.csv", header=TRUE)
glimpse(netflix_df)
```


## Summary Statistics
1. Calculate the mean and standard deviation of age and Netflix time

```{r}
netflix_df %>% summarise(mean_age = mean(age), sd_age = sd(age),
                         mean_time = mean(time), sd_time = sd(time))
```


### Scatterplot 

2. Create a scatterplot of Netflix time vs Ageand give a subjective impression of the relationship between the variables


```{r}

ggplot(netflix_df, aes(y=time, x=age)) + 
  geom_point() +
  labs(x = "Age (years)", y = "Netflix Time (mins)", 
       title = "Scatterplot")

```

The subjective impression is that lower Netflix times are observed among older individuals. We expect a negative correlation

### Correlation

3. Calculate and interpret the Pearson correlation coefficient between age and Netflix time

```{r}
netflix_df %>% select(time,age) %>% cor()
```

```{r}

ggplot(netflix_df, aes(y=time, x=age)) + 
  geom_point() +
  labs(x = "Age (years)", y = "Netflix Time (mins)", 
       title = "Correlation is -0.695")

```

The correlation coefficient $r=-0.96$ suggesting a very strong negative or inverse **linear** relationship between time on Netflix and age.

### Scatterplot of Netflix time vs Age with a Lowess smoother.

4. Create a scatterplot with a smoother and comment on whether this suggests a linear relationship.

```{r}

ggplot(netflix_df, aes(y=time, x=age)) + 
  geom_point() +
  geom_smooth() +
  labs(x = "Age (years)", y = "Netflix Time (mins)", 
       title = "Scatterplot with Lowess Smoother")

```

The smoother suggests that the overall trend is plausibly linear.

### Fit the model

5. Fit a regression model in `R`, include a summary of the model and create a scatterplot with the line of best fit

```{r}

netflix_model <- lm(time ~ age, data = netflix_df)

# regression tables

summary(netflix_model)

# tidier tables

get_regression_table(model = netflix_model)

```


```{r}

ggplot(netflix_df, aes(y=time, x=age)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Age (years)", y = "Netflix Time (mins)", 
       title = "Scatterplot with line of best fit")

```


## Interpret the results
6. What is equation of the line of best fit?

The regression equation is $Time=100.4 - 0.514Age$ suggesting that time is reduced in older people.

7. Interpret the intercept and slope estimates.

The information provided by the sample slope is that it is estimated that a 1 year increase in age will result in decreased Netflix Time of 0.514 minutes on average. It is questionable whether this is a meaningful decrease, e.g. people who are ten years older spend just 5.14 mins per day less on average!

The sample intercept states that those with age zero will have an average Netflix Time of 100.4 minutes. This makes no sense and highlights a very important feature of any regression line, namely that a regression line is only valid for the range of data collected.  The minimum age recorded was 17 and the predicted NEtflix Time for age would be 100.4-0.514*(17) = 91.7 minutes, which is much more plausible. In general, never try to extrapolate the regression equation outside the range of the explanatory variable!



